---
title: "Kernel Distance Metric Learning with kdml"
author: "Daniel Krasnov"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Kernel Distance Metric Learning with kdml}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  fig.retina = 1
)
library(kdml)
```

# Overview

Distance-based methods need a useful definition of dissimilarity. That is
particularly difficult when a data frame mixes continuous measurements,
nominal factors, and ordered factors. Kernel distance metric learning (KDML)
assigns a type-compatible kernel and bandwidth to each feature. The bandwidth
controls how strongly differences in that feature affect the result and can
therefore shrink irrelevant features toward uniform similarity.

For observations \(\mathbf{x}_i\) and \(\mathbf{x}_j\), `kdml` supports two
similarities. DKPS uses a product over continuous kernels and sums the nominal
and ordinal kernels; DKSS sums all feature kernels. Either similarity \(s\) is
converted to a squared kernel distance by

\[
d^2(\mathbf{x}_i,\mathbf{x}_j)
= s(\mathbf{x}_i,\mathbf{x}_i)
+ s(\mathbf{x}_j,\mathbf{x}_j)
- 2s(\mathbf{x}_i,\mathbf{x}_j).
\]

The traditional estimator chooses bandwidths by maximum similarity
cross-validation (MSCV). The MCMC interface treats the MSCV score as a loss
within a generalized Bayesian update, allowing uncertainty in both bandwidths
and feature-specific kernel choices. This vignette uses deliberately small
settings so it rebuilds quickly and reproducibly on the CPU. Increase the
warmup and retained draws for substantive analysis.

# Installation and optional CUDA support

The standard source and binary packages always provide the CPU backend.

```{r install-cpu, eval=FALSE}
install.packages("kdml")
library(kdml)
```

CUDA support is optional and is never required by package examples or
vignettes. A source installation can request it when a compatible NVIDIA GPU,
CUDA toolkit, and host compiler are available.

```{r install-cuda, eval=FALSE}
Sys.setenv(KDML_ENABLE_CUDA = "true")
install.packages("kdml", type = "source", INSTALL_opts = "--preclean")
```

The availability check is safe on CPU-only systems.

```{r inspect-cuda}
cuda_info()[c("compiled", "available", "device_count")]
```

# A small mixed-type example

`confactord()` generates continuous, nominal, and ordinal features together
with the true two-group membership. We use 36 observations and one feature of
each type.

```{r simulate-data}
set.seed(100)
simulated <- confactord(
  n = 36,
  popProb = c(0.5, 0.5),
  numMixVar = c(1, 1, 1),
  numMixVarOl = c(1, 1, 1),
  olVarType = c(0.20, 0.20, 0.20),
  catLevels = c(3, 4)
)

dat <- simulated$data
names(dat) <- c("continuous", "nominal", "ordinal")
true_group <- factor(simulated$class)
str(dat)
```

If bandwidths have already been chosen, `dkps()` and `dkss()` calculate a
distance directly. Numeric bandwidths follow data-frame column order. Here the
continuous bandwidth is on the data scale, while the categorical bandwidths
lie strictly between zero and one.

```{r fixed-distance}
fixed_bandwidths <- c(stats::sd(dat$continuous), 0.25, 0.25)
fixed <- dkps(dat, bw = fixed_bandwidths)
fixed$distances[1:4, 1:4]
```

For a full analysis, bandwidth point estimates can instead be obtained with
`mscv.dkps()` or `mscv.dkss()`. Those optimization routines are not run here
because their runtime depends strongly on the data and requested restarts.

```{r fit-mscv, eval=FALSE}
set.seed(100)
mscv_fit <- mscv.dkps(dat, nstart = 3)
distance_from_mscv <- dkps(dat, bw = mscv_fit$bw[, 1])
```

# Sampling kernels and bandwidths

The convenience functions `mcmc.dkps()` and `mcmc.dkss()` select the distance
definition. Multiple chains are important for convergence diagnostics. The
small run below exercises the complete workflow, but is much shorter than a
run intended for inference.

```{r fit-mcmc}
set.seed(101)
fit <- mcmc.dkps(
  dat,
  chains = 2,
  chain_threads = 1,
  warmup = 25,
  draws = 40,
  thin = 1,
  backend = "cpu",
  progress = FALSE
)
fit
```

The fit stores transformed bandwidths (`theta`), natural-scale bandwidths,
kernel names and codes, MSCV scores, log targets, proposal diagnostics, and a
draws array that follows the conventions of the `posterior` package. The `map`
component is the retained state
with the largest sampled log target.

```{r inspect-fit}
fit$map$kernel
fit$map$bandwidth
dim(posterior::as_draws_array(fit))
```

`dkps(fit)` reconstructs distances from the retained MAP state for a DKPS fit;
`dkss(fit)` provides the corresponding operation for a DKSS fit.

```{r mcmc-distance}
mcmc_distance <- dkps(fit)
mcmc_distance$distances[1:4, 1:4]
```

# Diagnostics

`summary()` combines transformed-bandwidth diagnostics with kernel occupancy,
kernel transitions, score summaries, and post-warmup acceptance rates. Short
demonstration chains will commonly trigger low-ESS or \(\widehat R\) warnings;
those warnings are a reason to run longer chains before interpreting a real
analysis.

```{r diagnostic-summaries}
fit_summary <- summary(fit, min_ess_per_chain = 10)
fit_summary$theta
fit_summary$convergence
fit_summary$kernel_occupancy
```

Trace and kernel-occupancy plots provide complementary views of continuous and
discrete mixing.

```{r trace-plots, fig.width=7, fig.height=4}
plot(fit, type = "trace")
```

```{r kernel-posteriors, fig.width=7, fig.height=4}
plot(fit, type = "kernel")
```

Other plot types are `"acf"`, `"diagnostics"`, `"score"`, `"bandwidth"`,
and `"acceptance"`. Features and chains can be selected by name or index.

# Using a learned distance

Because the returned values are squared kernel distances, take a square root
before supplying them to an algorithm that expects ordinary distances. As a
small illustration, we compare Ward clustering from the fixed state and the
sampled MAP state.

```{r clustering}
fixed_tree <- hclust(as.dist(sqrt(pmax(fixed$distances, 0))), method = "ward.D2")
mcmc_tree <- hclust(
  as.dist(sqrt(pmax(mcmc_distance$distances, 0))),
  method = "ward.D2"
)

fixed_cluster <- cutree(fixed_tree, k = 2)
mcmc_cluster <- cutree(mcmc_tree, k = 2)
table(true_group, fixed_cluster)
table(true_group, mcmc_cluster)
```

These compact results illustrate the mechanics, not an assessment of sampler
convergence or clustering performance. A real analysis should use more draws,
inspect all relevant diagnostics, and assess the stability of conclusions
across plausible retained states.

# References

Bissiri, P. G., Holmes, C. C., and Walker, S. G. (2016). A general framework
for updating belief distributions. *Journal of the Royal Statistical Society:
Series B*, 78(5), 1103--1130. <https://doi.org/10.1111/rssb.12158>.

Ghashti, J. S. (2024). *Similarity Maximization and Shrinkage Approach in
Kernel Metric Learning for Clustering Mixed-Type Data*. PhD thesis, University
of British Columbia. <https://doi.org/10.14288/1.0443975>.

Ghashti, J. S. and Thompson, J. R. J. (2025). Mixed-type distance shrinkage
and selection for clustering via kernel metric learning. *Journal of
Classification*, 42, 311--334.
<https://doi.org/10.1007/s00357-024-09493-z>.
