---
title: "Kernel Distance Metric Learning with kdml"
author: "John R. J. Thompson, Jesse S. Ghashti, and Daniel Krasnov"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Kernel Distance Metric Learning with kdml}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  cache = TRUE
)
library(kdml)
```

# Introduction

Distance-based clustering is widely used to group observations with mixed
numeric and categorical features, using a predefined metric to quantify
dissimilarity between data points. Many existing metrics for mixed-type data
either convert continuous attributes to categorical attributes, or vice versa,
and treat all variables as a single type, or calculate a distance for each
variable separately and then combine the results. These approaches can discard
information and allow irrelevant or poorly scaled variables to exert too much
influence on the resulting dissimilarities.

Kernel distance metric learning (KDML) addresses this challenge by assigning a kernel to each feature based on its data type. Continuous kernels measure numerical differences, nominal kernels capture agreement between categories, and ordinal kernels account for the inherent ordering of categories. Each feature is assigned a bandwidth parameter that determines the extent to which differences in that feature contribute to the overall similarity. 
As shown by Ghashti and Thompson (2025), KDML can be interpreted as a shrinkage method between maximum dissimilarity between all data
points to uniform dissimilarity across data points (Ghashti and Thompson, 2025).

For observations \(\mathbf{x}_i\) and \(\mathbf{x}_j\), let \(p_c\), \(p_u\),
and \(p_o\) denote the numbers of continuous, nominal, and ordinal features,
with \(p=p_c+p_u+p_o\). We define two similarity functions, the kernel product similarity

\[
s_{\mathrm{DKPS}}(\mathbf{x}_i,\mathbf{x}_j\mid\boldsymbol{\lambda})
=
\prod_{k=1}^{p_c}\frac{1}{\lambda_k}
K_k\left(\frac{x_{i,k}-x_{j,k}}{\lambda_k}\right)
+
\sum_{k=p_c+1}^{p_c+p_u}
L_k(x_{i,k},x_{j,k};\lambda_k)
+
\sum_{k=p_c+p_u+1}^{p}
\ell_k(x_{i,k},x_{j,k};\lambda_k),
\]

and the kernel summation similarity

\[
s_{\mathrm{DKSS}}(\mathbf{x}_i,\mathbf{x}_j\mid\boldsymbol{\lambda})
=
\sum_{k=1}^{p_c}
K_k\left(\frac{x_{i,k}-x_{j,k}}{\lambda_k}\right)
+
\sum_{k=p_c+1}^{p_c+p_u}
L_k(x_{i,k},x_{j,k};\lambda_k)
+
\sum_{k=p_c+p_u+1}^{p}
\ell_k(x_{i,k},x_{j,k};\lambda_k).
\]

Either similarity is converted to a squared kernel distance through

\[
d^2(\mathbf{x}_i,\mathbf{x}_j\mid\boldsymbol{\lambda})
=
s(\mathbf{x}_i,\mathbf{x}_i\mid\boldsymbol{\lambda})
+
s(\mathbf{x}_j,\mathbf{x}_j\mid\boldsymbol{\lambda})
-
2s(\mathbf{x}_i,\mathbf{x}_j\mid\boldsymbol{\lambda}).
\]

When using the product similarity, we refer to the metric as the distance using kernel product similarity (DKPS), and when using the summation similarity, we refer to the metric as the distance using kernel summation similarity (DKSS). 

The usual KDML estimator chooses the bandwidths by maximum similarity
cross-validation (MSCV). MSCV measures the
leave-one-out similarity of every observation to the remaining observations:

\[
S(\boldsymbol{\lambda};\mathbf{X})
=
\frac{1}{n}\sum_{i=1}^{n}
\log\left[
\frac{1}{n-1}
\sum_{\substack{j=1\\j\ne i}}^{n}
s(\mathbf{x}_i,\mathbf{x}_j\mid,\boldsymbol{\lambda})
\right].
\]

Maximizing the MSCV criterion estimates all feature bandwidths simultaneously, producing point estimates for the bandwidth parameters. If we are interested in quantifying uncertainty in both the kernels used and their bandwidths we may treat the negative MSCV criterion, \(-S\), as a loss function within an MCMC algorithm (Bissiri et al., 2016; Krasnov and Thompson, 2026). Let \(\boldsymbol{\theta}\) denote bandwidth parameters that have undergone a suitable transformation so they are unconstrained and \(m\) the collection of feature-specific kernels. The resulting posterior target is

\[
\pi(m,\boldsymbol{\theta}\mid\mathbf{X})
\propto
\exp\left\{\beta
S\bigl(m,g(\boldsymbol{\theta});\mathbf{X}\bigr)\right\}
p(\boldsymbol{\theta}\mid m)p(m),
\]

where \(g\) maps transformed parameters into the bandwidth domain for each kernel and \(\beta>0\) controls the relative influence of the MSCV criterion and the prior. The MCMC sampler updates each feature sequentially by jointly proposing a kernel and a random-walk update to its transformed bandwidth. The retained samples provide posterior distributions for the bandwidths, posterior probabilities for the kernel choices, and a sampled maximum a posteriori (MAP) estimate.

# Installation

The standard installation provides a CPU backend:

```{r install-cpu, eval=FALSE}
install.packages("kdml")
library(kdml)
```

When using the MCMC sampler we recommend using a CUDA device to  accelerate the repeated MSCV evaluations. To enable CUDA, you must have an NVIDIA GPU, the NVIDIA CUDA Toolkit, and `nvcc` installed. Then install the package using the following command:

```{r install-cuda, eval=FALSE}
Sys.setenv(KDML_ENABLE_CUDA = "true")
install.packages(
  "kdml",
  type = "source",
  INSTALL_opts = "--preclean"
)
```

To check if CUDA is available, run the following commands:

```{r verify-cuda, eval=TRUE}
library(kdml)

cuda_available()
```

# Example Analysis

`kdml` provides the `confactord()` function to generate mixed-type data. In this vignette, we generate 200 observations from two equally probable
clusters.

```{r simulate-data, fig.width=9, fig.height=9, out.width="100%", fig.align="center"}
set.seed(100)

simulated <- confactord(
  n = 200,
  popProb = c(0.5, 0.5),
  numMixVar = c(2, 2, 2),
  numMixVarOl = c(1, 1, 1),
  olVarType = c(0.15, 0.15, 0.15),
  catLevels = c(3, 4)
)

df <- simulated$data
names(df) <- c(
  "continuous_1", "continuous_2",
  "nominal_1", "nominal_2",
  "ordinal_1", "ordinal_2"
)

true_cluster <- factor(
  simulated$class,
  levels = c(1, 2),
  labels = c("Cluster 1", "Cluster 2")
)
cluster_colours <- c("#0072B2", "#D55E00")

pairs(
  data.matrix(df),
  pch = 19,
  cex = 0.5,
  col = grDevices::adjustcolor(
    cluster_colours[as.integer(true_cluster)],
    alpha.f = 0.6
  ),
  labels = gsub("_", " ", names(df)),
  main = "Simulated mixed-type data\nCluster 1 (blue), Cluster 2 (orange)"
)
```

Use `mscv.dkps()` to estimate the bandwidths, then pass them to `dkps()` to
obtain the pairwise distances:

```{r fit-mscv}
set.seed(100)

mscv_fit <- mscv.dkps(
  df = df,
  verbose = TRUE
)

dkps_mscv <- dkps(
  df = df,
  bw = mscv_fit$bw[, 1]
)

dkps_mscv$distances[1:5, 1:5]
```

Alternatively, estimate the kernels and bandwidths by MCMC, then pass the fit
to `dkps()` to obtain the pairwise distances:

```{r fit-mcmc}
set.seed(100)

fit_dkps_mcmc <- mcmc.dkps(
  df = df,
  chains = 4,
  chain_threads = 1,
  warmup = 5000,
  draws = 5000,
  thin = 5,
  backend = "cuda",
  progress = FALSE
)

dis_dkps_mcmc <- dkps(fit_dkps_mcmc)
dis_dkps_mcmc$distances[1:5, 1:5]
```

# Diagnostics

Before interpreting the MCMC results, it is important to assess convergence and sampling efficiency. The package provides trace, autocorrelation, and proposal-acceptance plots, along with rank-normalized \(\widehat{R}\) and bulk and tail effective sample sizes (ESS). Calling summary() on the fitted MCMC object provides a convenient overview of these diagnostics.

```{r diagnostic-summaries}
fit_summary <- summary(fit_dkps_mcmc)

fit_summary$theta
fit_summary$convergence
fit_summary$score
```

Trace plots can be accessed through the `plot()` with `type="trace"`. Here we expect to see that each MCMC chain has converged to the same stationary distribution.

```{r trace-plots, fig.width=8, fig.height=7}
diagnostic_features <- c(
  "continuous_1",
  "nominal_1",
  "ordinal_1"
)

plot(
  fit_dkps_mcmc,
  type = "trace",
  features = diagnostic_features
)
```

Next, autocorrelation plots can be obtained with `type="acf"`. An MCMC chain that has converged should have low autocorrelation.


```{r acf-plots, fig.width=8, fig.height=7}
plot(
  fit_dkps_mcmc,
  type = "acf",
  features = diagnostic_features
)
```

To view both \(\widehat{R}\) and the effective sample size (ESS) together, use `type = "diagnostics"`. The \(\widehat{R}\) statistic compares within-chain and between-chain variation to assess whether the MCMC chains have mixed and converged. Values close to 1 indicate good convergence, whereas values greater than 1 suggest the chains have not yet converged. The ESS measures the amount of independent information contained in the correlated MCMC draws, and can be interpreted as the equivalent number of independent samples represented by the chain.

```{r diagnostic-plots, fig.width=9, fig.height=5}
plot(fit_dkps_mcmc, type = "diagnostics")
```

We may also be interested in the trace plot of the score function which can be accessed via `type="score"`. Similar to other trace plots, we wish to see that each chain has converged to the same values.

```{r score-plots, fig.width=8, fig.height=6}
plot(fit_dkps_mcmc, type = "score")
```

Bandwidth density plots allow us to inspect the uncertainty in our estimates. These plots can be viewed with `type = "bandwidth"`. Continuous bandwidth densities are displayed on the natural-log
scale and nominal and ordinal bandwidths are plotted as is.

```{r bandwidth-posteriors, fig.width=8, fig.height=6}
plot(fit_dkps_mcmc, type = "bandwidth")
```

The MCMC algorithm provides posterior probabilities for each kernel choice. We can visualize these probabilities with `type = "kernel"`. First we look at the continuous kernels:

```{r kernel-posteriors-continuous, fig.width=10, fig.height=4.5}
plot(
  fit_dkps_mcmc,
  type = "kernel",
  features = c("continuous_1", "continuous_2"),
  args.legend = list(
    x = "topright",
    inset = c(-0.55, 0),
    bty = "n",
    cex = 0.7
  )
)
```

Next the nominal kernels:

```{r kernel-posteriors-nominal, fig.width=10, fig.height=4.5}
plot(
  fit_dkps_mcmc,
  type = "kernel",
  features = c("nominal_1", "nominal_2"),
  args.legend = list(
    x = "topright",
    inset = c(-0.55, 0),
    bty = "n",
    cex = 0.7
  )
)
```

Finally the ordinal kernels:

```{r kernel-posteriors-ordinal, fig.width=10, fig.height=4.5}
plot(
  fit_dkps_mcmc,
  type = "kernel",
  features = c("ordinal_1", "ordinal_2"),
  args.legend = list(
    x = "topright",
    inset = c(-0.55, 0),
    bty = "n",
    cex = 0.7
  )
)
```

Post-warmup acceptance rates are reported separately for kernel and
bandwidth proposals with `type = "acceptance"`. Very low rates can indicate proposals that are too
aggressive, while rates near one can indicate moves that are too small to
explore the posterior efficiently.

```{r acceptance-plot, fig.width=9, fig.height=5}
plot(fit_dkps_mcmc, type = "acceptance")
```


## Performing Clustering

We finish by comparing hierarchical clustering based on the standard MSCV
estimate and the MCMC MAP estimate.

```{r clustering}
hc_mscv <- hclust(
  as.dist(sqrt(pmax(dkps_mscv$distances, 0))),
  method = "ward.D2"
)
hc_mcmc <- hclust(
  as.dist(sqrt(pmax(dis_dkps_mcmc$distances, 0))),
  method = "ward.D2"
)

clusters_mscv <- cutree(hc_mscv, k = 2)
clusters_mcmc <- cutree(hc_mcmc, k = 2)
```

```{r clustering-plots, fig.width=9, fig.height=5}
plot(
  hc_mscv,
  labels = FALSE,
  hang = -1,
  main = "Ward Clustering using MSCV",
  xlab = ""
)
rect.hclust(hc_mscv, k = 2, border = cluster_colours)
plot(
  hc_mcmc,
  labels = FALSE,
  hang = -1,
  main = "Ward Clustering using MCMC",
  xlab = ""
)
rect.hclust(hc_mcmc, k = 2, border = cluster_colours)
```

We can see that each clustering method has correctly identified the two clusters. The following tables show the number of observations in each true cluster that were assigned to each estimated cluster.

```{r clustering-tables}
table(
  true_cluster = true_cluster,
  MSCV_cluster = clusters_mscv
)

table(
  true_cluster = true_cluster,
  MCMC_cluster = clusters_mcmc
)
```

**References**

[1] Bissiri, P.G., C.C. Holmes, and S.G. Walker (2016), "A General Framework
for Updating Belief Distributions", *Journal of the Royal Statistical Society:
Series B*, 78(5), 1103-1130.

[2] Ghashti, J.S. (2024), "Similarity Maximization and Shrinkage Approach in
Kernel Metric Learning for Clustering Mixed-Type Data", PhD thesis, University
of British Columbia.

[3] Ghashti, J.S. and J.R.J. Thompson (2025), "Mixed-Type Distance Shrinkage
and Selection for Clustering via Kernel Metric Learning", *Journal of
Classification*, 42, 311-334.
<https://doi.org/10.1007/s00357-024-09493-z>.

[4] Hayfield, T. and J.S. Racine (2008), "Nonparametric Econometrics: The
`np` Package", *Journal of Statistical Software*, 27(5).

[5] Krasnov, D. and J.R.J. Thompson (2026), "Bayesian Inference for Kernel
Metric Learning", unpublished manuscript.

[6] Li, Q. and J.S. Racine (2007), *Nonparametric Econometrics: Theory and
Practice*, Princeton University Press.

[7] Silverman, B.W. (1986), *Density Estimation for Statistics and Data
Analysis*, Chapman and Hall.

[8] Wang, M.C. and J. van Ryzin (1981), "A Class of Smooth Estimators for
Discrete Distributions", *Biometrika*, 68, 301-309.
